### 2015-08-05 conclusion.tex
Done

In this paper we presented \whoop, a new fully automatic approach for detecting all possible data races in device drivers. Compared to traditional data race detection techniques that are based on happens-before and typically attempt to explore as many thread interleavings as possible (and thus face code coverage and scalability issues), \textsc{Whoop} uses over-approximation and symbolic pairwise lockset analysis, which has the potential to scale well. Exploiting the race-freedom guarantees provided by \whoop, we showed that we can achieve a sound partial-order reduction that can significantly accelerate \corral, a state-of-the-art concurrent bug-finder. Combining \whoop and \corral, we analyzed \sizeOfBenchmarks drivers from the Linux 4.0 kernel, showing that our technique can analyze drivers 1.5--20$\times$ faster than standalone \corral.

>>>>

In this paper we presented \whoop, a new automated approach for detecting all possible data races in device drivers. Compared to traditional data race detection techniques that are based on happens-before and typically attempt to explore as many thread interleavings as possible (and thus face code coverage and scalability issues), \textsc{Whoop} uses over-approximation and symbolic pairwise lockset analysis, which scales well. Exploiting the race-freedom guarantees provided by \whoop, we showed that we can achieve a sound partial-order reduction that can significantly accelerate \corral, a state-of-the-art concurrency bug-finder. Combining \whoop and \corral, we analyzed \sizeOfBenchmarks drivers from the Linux 4.0 kernel, showing that our technique can analyze drivers 1.5--20$\times$ faster than standalone \corral.

### 2015-08-05 evaluation.tex
Done

\begin{figure}
\centering
\includegraphics[width=.99\linewidth]{experiments/figures/yieldmr_vs_yieldall.pdf}
%\vspace{-2mm}
\caption{Scatter plot showing the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation. The symbols $+$, $\circ$ and $\times$, represent a context-switch bound of 2, 5 and 9, respectively.}
\label{fig:plot}
%\vspace{-2mm}
\end{figure}

@@

\noindent\textbf{Benchmarks }
%
We evaluate our methodology against \sizeOfBenchmarks drivers from the Linux 4.0 kernel. We chose non-trivial drivers of several types: block; char; ethernet; near field communication (nfc); universal serial bus (usb); and watchdog. For all these drivers, we had to understand their environment and manually model it; this required approximately two months of work.

\noindent\textbf{Race-Checking }
%
Table~\ref{tab:stats} presents statistics for our benchmarks: lines of code (LoC); number of possible entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of data races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found).\footnote{The number of racy memory regions can be less than the number of races found by \corral, because \whoop might find that a memory region is racy, but the same memory region might race in many points in the program.} Using a higher csb did not uncover any further races; this might mean that all races in our benchmarks can manifest with a csb of $\leq$ 2, or that \corral hit its recursion depth bound before discovering a deeper bug.

@@

As an example of the second category, we found the following race in the generic\_nvram driver (see Fig.~\ref{fig:data_race_example}): the \texttt{llseek} entry point accesses the file offset \texttt{file->f\_pos} without a lock (\texttt{file} is passed as a parameter to \texttt{llseek}). This causes a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.  We observed that another char driver, using the same APIs, \emph{does} use a mutex to protect the offset access in its \texttt{llseek} entry point, leading us to suspect that the race we found in generic\_nvram is not benign.  We have filed a bug report in the Linux kernel bug tracker and are awaiting a response.

\noindent\textbf{Accelerating \corral }
%
Table~\ref{tab:results} presents runtime results from using \whoop, \corral and \whoop + \corral to analyze our benchmarks, while Fig.~\ref{fig:plot} plots the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation. All reported times are in seconds and averaged over three runs. \corral was configured with a time budget of 10 hours (T.O. denotes timeout) and a csb of 2, 5 and 9. Standalone \corral uses \yieldall, which instruments context-switches (i.e. \texttt{yield} statements) after all visible operations, while \whoop + \corral uses \yieldcoarse and \yieldmr, which instrument context-switches in a more fine-grained fashion (see \S\ref{corral}). The table also shows the number of context-switches per instrumentation (\#Y).

\whoop uses over-approximation to scale and, as expected, executes significantly faster than \corral in all our benchmarks. For example, \corral times out in all configurations (with and without \whoop) when trying to analyze the r8169 ethernet driver, while \whoop manages to analyze the driver in 144.5 seconds. We believe that the reason behind this is that r8169 has deeply-nested recursion in some of its entry points, which might cause \corral to get stuck. This is not an issue for \whoop, which uses procedure summarization. This shows that \whoop has value as a stand-alone analyzer.

Using the race-freedom guarantees from \whoop, we managed to significantly accelerate \corral in most of our benchmarks; the best results were achieved using \yieldmr. Fig.~\ref{fig:plot} shows that most speedups using \yieldmr are between 1.5$\times$ and 10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup of 12$\times$, and 20$\times$ respectively, with a csb of 9. We noticed that a higher csb typically results into greater speedups when exploiting \whoop. This is expected, as a higher csb translates into a larger sequentialized program, and thus \whoop has more opportunities to reduce this sequentialization.
%
However, there are some cases where \whoop might slow down \corral. We noticed this in the sx8 driver: \whoop verified 46 out of 47 memory regions, but did not fully verify any of the pairs; \corral, on the other hand, analyzed the only two pairs of the driver in just 21.4 seconds (csb of 9). We believe that in this case the overhead of running \whoop outweighed the benefits of using \yieldmr.

\noindent\textbf{Other Tools }
%
We tried to compare \whoop with other similar approaches (see \S\ref{related}). However, we found this to be hard in practice: we downloaded Locksmith~\cite{pratikakis2006locksmith}, but could not get it to work with the 4.0 Linux drivers (the tool was last updated in 2007); we also could not find source code or binaries for \cite{kahlon2007fast, kahlon2009semantic, das2015section}.

>>>>

\begin{figure}
\centering
\includegraphics[width=.99\linewidth]{experiments/figures/yieldmr_vs_yieldall.pdf}
%\vspace{-2mm}
\caption{Scatter plot showing the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation. The symbols $+$, $\circ$, and $\times$ represent a context-switch bound of 2, 5, and 9, respectively.}
\label{fig:plot}
%\vspace{-2mm}
\end{figure}

@@

\noindent\textbf{Benchmarks }
%
We evaluate our methodology against \sizeOfBenchmarks drivers from the Linux 4.0 kernel. We chose non-trivial drivers of several types: block, char, ethernet, near field communication (nfc), universal serial bus (usb), and watchdog. For all these drivers, we had to understand their environment and manually model it; this required approximately two months of work.

\noindent\textbf{Race-Checking }
%
Table~\ref{tab:stats} presents statistics for our benchmarks: lines of code (LoC); number of possible entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of data races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found).\footnote{The number of racy memory regions can be less than the number of races found by \corral: \whoop might find that a memory region is racy, but the same memory region might race in several program statements.} Using a higher csb did not uncover any further races; this might mean that all races in our benchmarks can manifest with a csb of 2, or that \corral hit its recursion depth bound before discovering a deeper bug.

@@

As an example of the second category, we found the following race in the generic\_nvram driver (see Fig.~\ref{fig:data_race_example}): the \texttt{llseek} entry point accesses the file offset \texttt{file->f\_pos} without holding a lock (\texttt{file} is passed as a parameter to \texttt{llseek}). This causes a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.  We observed that another char driver, using the same APIs, \emph{does} use a mutex to protect the offset access in its \texttt{llseek} entry point, leading us to suspect that the race we found in generic\_nvram is not benign.  We have filed a bug report in the Linux kernel bug tracker and are awaiting a response.

\noindent\textbf{Accelerating \corral }
%
Table~\ref{tab:results} presents runtime results from using \whoop, \corral, and \whoop + \corral to analyze our benchmarks, while Fig.~\ref{fig:plot} plots the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation. All reported times are in seconds and averaged over three runs. \corral was configured with a time budget of 10 hours (T.O. denotes timeout) and a csb of 2, 5, and 9. Standalone \corral uses \yieldall, which instruments context-switches (i.e. \texttt{yield} statements) after all visible operations, while \whoop + \corral uses \yieldcoarse and \yieldmr, which instrument context-switches in a more fine-grained fashion (see \S\ref{corral}). The table also shows the number of context-switches per instrumentation (\#Y).

\whoop uses over-approximation to scale and, as expected, executes significantly faster than \corral in all our benchmarks. For example, \corral times out in all configurations (with and without \whoop) when trying to analyze the r8169 ethernet driver, while \whoop manages to analyze the driver in 144.5 seconds. We believe that the reason behind this is that r8169 has deeply-nested recursion in some of its entry points, which puts burden on \corral's stratified inlining. This is not an issue for \whoop, which uses procedure summarization. This shows that \whoop has value as a stand-alone analyzer.

Using the race-freedom guarantees from \whoop, we managed to significantly accelerate \corral in most of our benchmarks; the best results were achieved using \yieldmr. Fig.~\ref{fig:plot} shows that most speedups using \yieldmr are between 1.5$\times$ and 10$\times$; in ssu100 and pc8736x\_gpio with a csb of 9 we achieved a speedup of 12$\times$ and 20$\times$, respectively. We noticed that a higher csb typically results in greater speedups when exploiting \whoop. This is expected as complexity grows exponentially with csb, and hence \whoop's reduction helps more at a higher csb.
%
However, there are cases where \whoop might slow down \corral. We noticed this in the sx8 driver: \whoop verified 46 out of 47 memory regions, but did not fully verify any of the pairs; \corral, on the other hand, analyzed the only two pairs of the driver in just 21.4 seconds (csb of 9). We believe that in this case the overhead of running \whoop outweighed the benefits of using \yieldmr.

\noindent\textbf{Other Tools }
%
We tried to compare \whoop with other similar approaches (see \S\ref{related}). However, we found this to be hard in practice: we downloaded Locksmith~\cite{pratikakis2006locksmith}, but could not get it to work with the 4.0 Linux drivers (the tool was last updated in 2007); we also could not find source code or binaries of other tools~\cite{kahlon2007fast, kahlon2009semantic, das2015section}.

### 2015-08-05 related.tex
Done

Static race analysis is a promising alternative to dynamic techniques, which restrict analysis to the schedule chosen by a (possibly controlled) scheduler, providing limited coverage~\cite{musuvathi2008finding}. Warlock~\cite{sterling1993warlock} and LockLint~\cite{oracle2010locklint} are notable static race analyzers.  In comparison to \whoop, these tools rely heavily on user annotations.
%
Most related to our lockset analysis are the static lockset analyzers RELAY~\cite{voung2007relay} and Locksmith~\cite{pratikakis2006locksmith}. Both tools, though, have several limitations. RELAY found 5022 warnings when analyzing the Linux kernel, with only 25 of them being true data races. To tackle this issue, RELAY employs unsound post-analysis filters and, hence, can also filter out real bugs. Locksmith was successfully applied in several small Pthreads applications and 7 medium-sized Linux device drivers, but the authors reported that the tool was unable to run on several large programs, hinting at its limited scalability. \whoop aims to achieve scalability \emph{and} precision: the first, via novel symbolic pairwise lockset analysis; and the second, by accelerating \corral, an industrial-strength precise bug-finder.

Prior works on static race checking also include: \cite{choi2002efficient}, which combines static analysis and runtime access caching to speed up dynamic race detection; \cite{kahlon2007fast}, which uses a divide-and-conquer algorithm that partitions all pointers of a program that do not alias in disjoint sets to achieve scalability; \cite{kahlon2009semantic}, which uses abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false race warnings; and \cite{das2015section}, which employs inter-procedural alias analysis and verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection. Our tool, \whoop, uses symbolic lockset analysis, which involves generating verification conditions and discharging them to a theorem prover, and then uses \corral to filter out false races.

>>>>

Static race analysis is a promising alternative to dynamic techniques, which restrict analysis to the schedule chosen by a (possibly controlled) scheduler, providing limited coverage~\cite{musuvathi2008finding}. Warlock~\cite{sterling1993warlock} and LockLint~\cite{oracle2010locklint} are notable static race analyzers.  In comparison to \whoop, these tools rely heavily on user annotations.
%
Most related to our lockset analysis are the static lockset analyzers RELAY~\cite{voung2007relay} and Locksmith~\cite{pratikakis2006locksmith}. Both tools, though, have several limitations. RELAY found 5022 warnings when analyzing the Linux kernel, with only 25 of them being true data races. To tackle this issue, RELAY employs unsound post-analysis filters and, hence, can also filter out real bugs. Locksmith was successfully applied in several small Pthreads applications and 7 medium-sized Linux device drivers, but the authors reported that the tool was unable to run on several large programs, hinting at its limited scalability. \whoop aims to achieve scalability \emph{and} precision: the first via novel symbolic pairwise lockset analysis, and the second by accelerating \corral, an industrial-strength precise bug-finder.

Prior works on static race checking also include the following. Choi et al.~\cite{choi2002efficient} combine static analysis and runtime access caching to speed up dynamic race detection. Kahlon et al.~\cite{kahlon2007fast} use a divide-and-conquer algorithm that partitions all pointers of a program that do not alias in disjoint sets to achieve scalability; more recently, they used abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false race warnings~\cite{kahlon2009semantic}. Das et al.~\cite{das2015section} employ inter-procedural alias analysis and verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection. \whoop uses symbolic lockset analysis, which involves generating verification conditions and discharging them to a theorem prover, and then employs \corral to filter out false races.
